---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Load some useful packages.
```{r}
library(lubridate)
library(dplyr)
library(data.table)
library(lattice)
```

## Loading and preprocessing the data

First we need to load the fitness tracker data from the website. 

```{r}

if (!file.exists('data')) {
        dir.create('data')
}

fileUrl <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
temp = tempfile()
download.file(fileUrl, destfile = ".\\data\\steps.zip")
fitness <- read.csv(unz(".\\data\\steps.zip", 'activity.csv'), stringsAsFactors = FALSE)

head(fitness)
```


## What is mean total number of steps taken per day?

Now we compute the total number of steps taken each day and plot a histogram of it.
```{r}
total_steps_per_day <- fitness %>% group_by(date) %>% summarize(steps = sum(steps, na.rm = TRUE))

hist(total_steps_per_day$steps, breaks = 10, xlab = 'Total number of steps per day', main = '', col = 'blue')
```

Next we calculate the mean and median of the total number of steps taken per day:
```{r}
mean_steps_per_day <- mean(total_steps_per_day$steps, na.rm = T)
median_steps_per_day <- median(total_steps_per_day$steps, na.rm = T)

print(paste('The mean number of steps per day is', round(mean_steps_per_day)))

print(paste('The median number of steps per day is', round(median_steps_per_day)))

```


## What is the average daily activity pattern?

We group the data by interval instead of date in order to compute the average number of steps in each interval across all days, and plot a time series plot of the acitivity pattern.

```{r}
mean_steps_by_interval <- fitness %>% group_by(interval) %>% summarize(steps = mean(steps, na.rm = TRUE))
plot(mean_steps_by_interval$interval, mean_steps_by_interval$steps, type = 'l', xlab = 'Interval', ylab = 'Number of steps')

```

Next, we calculate the interval with the maximum number of steps on average:

```{r}
truth <- max(mean_steps_by_interval$steps)
max_steps <- mean_steps_by_interval$interval[mean_steps_by_interval$steps == truth]

print(paste('The interval with the maximum number of steps on average is',max_steps))
```

## Imputing missing values

What is the total number of missing values?
```{r}
missing_data <- is.na(fitness$steps)
total_missing_data <- sum(missing_data)
print(paste('The total number of missing values is', total_missing_data))
```

In order to devise a strategy how to impute the missing data values we first need to find out if there is some structure to which data is missing. For example, is data missing for a complete day, is data missing for one interval for all days, etc.

```{r}
print('Table of intervals with missing data.')
table(fitness$interval[missing_data])
print('Table of days with missing data.')
table(fitness$date[missing_data])
print(paste('Number of measurements for one day', length(unique(fitness$interval))))


total_steps_per_day <- fitness %>% group_by(date) %>% summarize(steps = sum(steps, na.rm = F))

print(paste('Number of days that have no recorded tracking data',sum(is.na(total_steps_per_day$steps))))
```

We can see that 8 whole days of measurements are missing. So instead of imputing missing data on a per day basis the most promising approach for now is to use the intervals from other days to impute the missing data.

```{r}

fitness_imputed <- copy(fitness)
missing_data_ind <- which(missing_data)
for (i in missing_data_ind) {fitness_imputed$steps[i] <- median(fitness_imputed$steps[fitness_imputed$interval == fitness_imputed$interval[i]], na.rm = TRUE)}

head(fitness_imputed)
```

Let's see how the data has changed after the missing values have been replaced with the median value of the other intervals.

```{r}
total_steps_per_day_imputed <- fitness_imputed %>% group_by(date) %>% summarize(steps = sum(steps, na.rm = TRUE))

hist(total_steps_per_day_imputed$steps, breaks = 10, xlab = 'Total number of steps per day', main = '', col = 'blue')

mean_steps_per_day_imputed <- mean(total_steps_per_day_imputed$steps)
median_steps_per_day_imputed <- median(total_steps_per_day_imputed$steps)

print(paste('The mean number of steps per day after imputing missing values is', round(mean_steps_per_day_imputed)))

print(paste('The median number of steps per day after imputing missing values is', round(median_steps_per_day_imputed)))
```

We see that the mean number of steps has increased while the median number of steps has remained unchanged.


## Are there differences in activity patterns between weekdays and weekends?

In the last part we will check if there are differences during weekdays and weekends for the activity patterns.

```{r}
workday <- c('Monday','Tuesday','Wednesday', 'Thursday', 'Friday')
fitness_imputed$weekend <- ifelse(weekdays(ymd(fitness_imputed$date)) %in% workday,'Weekday','Weekend')
fitness_imputed$weekend <- factor(fitness_imputed$weekend, levels = c('Weekday', 'Weekend'))

head(fitness_imputed)
```

Plot the average activity pattern for weekdays and weekends.
```{r}
fitness_imputed_weekday <- fitness_imputed %>% group_by(weekend, interval) %>% summarize(steps = mean(steps, na.rm = TRUE))

xyplot(steps ~ interval | weekend, data = fitness_imputed_weekday, layout = c(1,2), xlab = 'Interval', ylab = 'Number of Steps', type = 'l')

```

What we could do next:
As the activity pattern for weekdays and weekends is noticeably different we should test how the results change if we impute missing data separately for these two. 